Prefetching SPARQL query cacher

### 2016-07-29 dissertation/cache-proxy.tex
more redis
detail the analyser in a section of its own

@@

\subsubsection{Analysis and Prefetching}\label{sec:analpre}

Analysers and prefetchers run asynchronously with query
evaluation. This is achieved by an addition to the proxy that sends a
query to a persistent analyser script that subscribes to a channel on
a Redis\footnote{See \url{http://redis.io/}} data structure store. The
Redis data structure store is a mostly-memory ``NoSQL'' database for
storing primitive data structures identified by a key.

The analyser script can be configured to use any number of analysers.
Two such analysers have been implemented, for now constrained to
single triple patterns. One will count the predicates in consecutive
queries it sees, and publish triple patterns to another Redis channel
with the predicate if the count exceeds a certain configurable
limit. This is an attempt to capture and cache popular parts of
queries.

The second analyser will rerun the query planner on the incoming query
while simulating that the results of a certain triple pattern was in
the cache. If the cost reduction of the best plan is larger than a
configurable limit, it will publish that triple pattern to the same
Redis channel as above.

@@

The query planner is written as completely stand-alone modules, based
on the facilities in the Attean framework, but without accessing
Attean internals. The system has in fact three query planners:
\pmodule{AtteanX::QueryPlanner::Cache},
\pmodule{AtteanX::QueryPlanner::Cache::LDF} and
\pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}. Each is an
extension of the previous. The first is written to use a remote
endpoint in combination with a local cache. The second extends this to
also be able to generate and use TPF plans. The
final is a query planner for the analyser in Section~\ref{sec:analpre}
that reruns the query planning with all triple patterns in a simulated
cache.

@@

The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
Protocol client that implements Attean APIs. The store implementation
itself composes the \mbox{TripleStore} API and implements methods to retrieve
the results of a single triple pattern and exact cardinality for a
given triple pattern, by using an aggregate query. In addition, it can
generate plans for Basic Graph Pattern algebra objects that have more
than one triple pattern. If that is the case, it will return an
instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
also defined by the module. If it is not the case, the module also has
an implementation of the \pcode{access\_plan} method in a
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
composed by a query planner to provide a single triple pattern
\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
model implementation that composes the Model and \mbox{CostPlanner}
APIs. This contains a \pcode{cost\_for\_plan} implementation that will
provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
proportional to the number of triple patterns in the plan, but
penalises plans that do not have triple patterns that are connected
through a variable (i.e. will cause a Cartesian join) with a factor
10. See Section~\ref{sec:costheuristics} for details on the cost
heuristics.

>>>>

\subsection{Analysis for prefetching}\ref{sec:analpre}

The objective of the analyser in Figure~\ref{fig:arch} is to predict
what remote data is most useful to answer future queries. I haven't
attempted to formalise the notion of what a useful cache entry is, nor
have I attempted to exploit the concept of profitable query patterns
of \cite{papailiou2015graph}, mostly because this work came at a time
where my work was already near the deadline. I opted for a simple
approach to consider only individual triple patterns.

With this constraint, the analysers have been implemented with two
different assumptions of usefulness: In the first case, I try to
capture frequently occurring bound predicates. This is due to the
observation that predicates are often bound, and so predicates that
are common will occur in the future, and therefore, prefetching and
caching results of a triple pattern where the predicate is bound to a
frequently occurring is likely to be beneficial. This is done by
counting the occurrences of predicates, and when the number exceeds a
certain threshold, the prefetcher is instructed to fetch the data.

The second analyser works along a different axis: It tries to find the
triple pattern that, if cached, would provide the greatest cost
benefit. To do so, it will rerun the query planner on the incoming
query while simulating that the results of a certain triple pattern
was in the cache. The simulation will run for \emph{all} triple
patterns that do not already have a cached result. If the cost
reduction of the best plan is larger than a certain limit, the
prefetcher is invoked.

@@

\subsubsection{Analysis and Prefetching}\label{sec:analimpl}

Analysers and prefetchers run asynchronously with query
evaluation. This is achieved by an addition to the proxy that sends a
query to a persistent analyser script that subscribes to a channel on
a Redis\footnote{See \url{http://redis.io/}} data structure store. The
Redis data structure store is a mostly-memory ``NoSQL'' database for
storing primitive data structures identified by a key. It also
features a publish-subscribe system.

The analyser script can be configured to use any number of analysers.
Two such analysers have been implemented as according to the ideas in
Section~\ref{sec:analpre}. Any further analysers that operate on
single triple patterns are trivial to implement, whereas analysers for
more advanced query patterns would require new plan classes as the
current cache system relies on using the \pcode{access\_plans} method,
which is applicable only to single triple/quad patterns. Any triple
patterns found by the analysers is published by them to another Redis
channel for the prefetcher to retrieve.

@@

The query planner is written as completely stand-alone modules, based
on the facilities in the Attean framework, but without accessing
Attean internals. The system has in fact three query planners:
\pmodule{AtteanX::QueryPlanner::Cache},
\pmodule{AtteanX::QueryPlanner::Cache::LDF} and
\pmodule{AtteanX::Query::Cache::Analyzer::QueryPlanner}. Each is an
extension of the previous. The first is written to use a remote
endpoint in combination with a local cache. The second extends this to
also be able to generate and use TPF plans. The
final is a query planner for the analyser in Section~\ref{sec:analimpl}
that reruns the query planning with all triple patterns in a simulated
cache.

@@

The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
Protocol client that implements Attean APIs. The store implementation
itself composes the \mbox{TripleStore} API and implements methods to retrieve
the results of a single triple pattern and exact cardinality for a
given triple pattern, by using an aggregate query. In addition, it can
generate plans for Basic Graph Pattern algebra objects that have more
than one triple pattern. If that is the case, it will return an
instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
also defined by the module. If it is not the case, the module also has
an implementation of the \pcode{access\_plans} method in a
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
composed by a query planner to provide a single triple pattern
\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
model implementation that composes the Model and \mbox{CostPlanner}
APIs. This contains a \pcode{cost\_for\_plan} implementation that will
provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
proportional to the number of triple patterns in the plan, but
penalises plans that do not have triple patterns that are connected
through a variable (i.e. will cause a Cartesian join) with a factor
10. See Section~\ref{sec:costheuristics} for details on the cost
heuristics.
